---
title: "Analysis code Rmarkdown"
author: "Lauke Stoel"
date: "26/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load packages, include=FALSE}
library(dplyr)
library(tidyverse)
```

Set working directory and load data
```{r}
setwd("~/Universiteit Utrecht/2020-2021/Semester 1/Survey Data Analysis/COVID Assignment")
dat <- readRDS("covid cases waste + tests.RDS")
```


### DATA CLEANING

Rename column names to include a code for the level of measurement: 
SR = Security Region
MN = Municipality
TP = Treatment Plant

```{r}
dat <- dat %>% rename(SR_code = code, 
                      SR_name = naam,
                      MN_name = Municipality_name, 
                      MN_code = Municipality_code, 
                      MN_province = Province,
                      MN_total.reported = Total_reported, 
                      MN_hospital.admisstions = Hospital_admission, 
                      MN_deceased = Deceased, 
                      MN_pop = inhabitants, 
                      MN_popperkm2 = perkm.2, 
                      MN_sizekm2 = size,
                      TP_code = RWZI_AWZI_code, 
                      TP_name = RWZI_AWZI_name,
                      TP_Xco = X_coordinate,
                      TP_Yco = Y_coordinate, 
                      TP_postal.code = Postal_code, 
                      TP_security.region = Security_region_name, 
                      TP_percentage.in.SR = Percentage_in_security_region,
                      TP_RNA.per.ml = RNA_per_ml, 
                      TP_RNA.flow.per.100000 = RNA_flow_per_100000, 
                      TP_representative.measurement = 
                      Representative_measurement)
```

Re-order the data such that those of the same level of measurement are put together. The date_of_report variable has been left out - it's the same as newdate and newdate has a more convenient notation. 
```{r}
dat <- dat %>% select(newdate, SR_code, SR_name, MN_code, MN_name, MN_code:TP_RNA.per.ml, TP_RNA.flow.per.100000, TP_representative.measurement)
```

Structure of the dataset: 
```{r}
str(dat)
```
Some variables are reported as character strings, that should be numeric variables - such as MN_sizekm2, TP_percentage.in.SR and TP_RNA.flow.per.100000. This is because the Dutch notation for decimal numbers is comma's, where R would expect dots, and vice versa. Other variables are noted in the wrong order of size because of this. 
```{r}
#replace comma's by dots for all variables with decimals
dat$MN_sizekm2 <- gsub(",", ".", dat$MN_sizekm2)
dat$MN_popperkm2 <- gsub(",", ".", dat$MN_popperkm2)
dat$TP_percentage.in.SR <- gsub(",", ".", dat$TP_percentage.in.SR)
dat$TP_RNA.flow.per.100000 <- gsub(",", ".", dat$TP_RNA.flow.per.100000)

#multiply by 1000 to get rid of the "." in the population variable
dat <- dat %>% mutate(MN_pop = MN_pop*1000)

#redefine variables as numeric:
dat$TP_Yco <- as.numeric(dat$TP_Yco)
dat$MN_popperkm2 <- as.numeric(dat$MN_popperkm2)
dat$MN_sizekm2 <- as.numeric(dat$MN_sizekm2)
dat$TP_RNA.flow.per.100000 <- as.numeric(dat$TP_RNA.flow.per.100000)
dat$TP_percentage.in.SR <- as.numeric(dat$TP_percentage.in.SR)

#check the structure again
str(dat)
```


### CONTENT OF THE ASSIGNMENT

#Step 1: New cases for each day + extra: difference in new cases per day

We are only considering dates from the 7th of September onward, because from that date, all sewage treatment plants joined in reporting data, and the type of measurement changes from RNA per milliliter to RNA flow per 100,000 inhabitants in the municipality. 

```{r}
#create dataframe with only the observations from 7th of September onward
all_dates <- unique(dat$newdate) #243 recorded dates, starting March 13th
used_dates <- dat %>% filter(newdate %in% all_dates[-c(1:178)]) #dates from 7th of Sept
```

Calculating the total number of cases for each day, for any selection of this dataset we enter into the formula. Outputs a data frame and a plot of the totals. 
```{r}
#this function calculates the total of reported cases on one day, across all municipalities:
totalCalculator <- function(sewer_data) {
  dates <- unique(sewer_data$newdate) #create a list of unique dates in the dataset 
  totals <- numeric(length(dates)) #make an object with the length of the number of dates
  
  #subtract nr of reported cases of the day before each day to get the difference
  for (i in 1:(length(totals))) {
    step1 <- filter(sewer_data, newdate == dates[i]) #select date we want to calculate the nr of new cases for
    
    #create data frame of the  dates, and remove duplicates (total reported per municipalities are the same across duplicates)
    df <- step1[!duplicated(step1[,"MN_name"]),] 
    
    #calculate the sum of all new reported cases per municipality and store in object
    sums <- sum(df$MN_total.reported, na.rm = TRUE)
    totals[i] <- sums
  }
  
  #create a data frame with all the outcomes, matched with each date
  totals_frame <- as.data.frame(cbind(dates, totals))
  colnames(totals_frame) <- c("Date", "Total number of reported infections") #rename columns
  totals_frame$`Total number of reported infections` <- as.numeric(totals_frame$`Total number of reported infections`) #store the totals as numeric values.
  
  #create a plot of the totals over time
  plot <- barplot(totals_frame$`Total number of reported infections`, main = "Total number of reported infections per day", xlab = "Date", ylab = "Total number of reported infections", ylim = c(0,350000), names.arg = as.Date(dates))
  
  #Output a data frame with the total of new cases per day
  output = list(totals_frame, plot)
  return(output)
}

totalCalculator(used_dates)

```


```{r}
#this function calculates the difference in reported cases between one day and the day before that:
differenceCalculator <- function(sewer_data) {
  dates <- unique(sewer_data$newdate) #create a list of unique dates in the dataset
  differences <- numeric(length(dates)-1) #make an object with the length of one fewer than the number of        dates. Why one fewer? We cannot calculate the number of new reported cases of the first day, because the number of cases from the preceding day is unknown.  
  
  #subtract nr of reported cases of the day before each day to get the difference
  for (i in 1:(length(differences))) {
    step1 <- filter(sewer_data, newdate == dates[i]) #select date of reference
    step2 <- filter(sewer_data, newdate == dates[i+1]) #select date we want to calculate the nr of new cases for
    
    #create data frames of both sets of dates, and remove duplicates (total reported per municipalities are the same across duplicates)
    df1 <- step1[!duplicated(step1[,"MN_name"]),] 
    df2 <- step2[!duplicated(step2[,"MN_name"]),]
    
    #calculate the difference between the two dataframes for each set of dates and store in object
    diff <- (sum(df2$MN_total.reported, na.rm = TRUE) - sum(df1$MN_total.reported, na.rm = TRUE))
    differences[i] <- diff
  }
  
  #create a data frame with all the outcomes, matched with each date
  diff_frame <- as.data.frame(cbind(dates[-1], differences))
  colnames(diff_frame) <- c("Date", "Difference in reported infections") #rename columns
  diff_frame$`Difference in reported infections` <- as.numeric(diff_frame$`Difference in reported infections`) #store the differences as numeric values.
  
  
  #create a plot of the totals over time
  plot <- barplot(diff_frame$`Difference in reported infections`, main = "Difference in reported infections compared to the previous day", xlab = "Date", ylab = "Difference in reported infections", names.arg = as.Date(dates[-1]))
  
  #output a data frame with new cases per day
  output = list(diff_frame, plot)
  return(output)
}

differenceCalculator(used_dates)
```

#Step 2: mean number of infections per 100.000 inhabitants per day
```{r}
#in the data preparation, we already multplied the municipality population by 1000,
#so no need to do that again.
#however, we do need make a new variable containing the number of cases per 100.000 people
#to do that, we multiple the total number of reported cases by 100.000 divided by the municipality population and give it an informative name.
used_dates$cases_per_100000 <- used_dates$MN_total.reported * (100000/used_dates$MN_pop)

#make a function that calculates the mean number of infection per 100.000 inhabitants per day.
#so, this is over the Netherlands as a whole.
differenceCalculator2 <- function(used_dates) {
  dates <- unique(used_dates$newdate) #create a list of dates
  differences <- numeric(length(dates)-1) #make an object with the length of one fewer than the number of dates. Why one fewer? We cannot calculate the number of new reported cases of the first day, because the number of cases from the preceding day is unknown.
  for (i in 1:(length(differences))) {
    step1 <- filter(used_dates, newdate == dates[i]) #select date of reference
    step2 <- filter(used_dates, newdate == dates[i+1]) #select date we want to calculate the nr of new cases for
    
    #create data frames of both sets of dates, and remove duplicates (total reported per municipalities are the same across duplicates)
    df1 <- step1[!duplicated(step1[,"MN_name"]),]
    df2 <- step2[!duplicated(step2[,"MN_name"]),]
    
    #calculate the differences by subtracting the previous day from the one the for loop is on then
    diff <- (mean(df2$cases_per_100000, na.rm = TRUE) - mean(df1$cases_per_100000, na.rm = TRUE))
    #and store it in our differences vector
    differences[i] <- diff
  }
  
  #create a dataframe with the differences by the respecitve dates
  diff_frame <- as.data.frame(cbind(dates[-1], differences))
  #give the columns informative names
  colnames(diff_frame) <- c("Date", "Increase in reported infections per 100.000")
  #and make the increase numeric
  diff_frame$`Increase in reported infections per 100.000` <- as.numeric(diff_frame$`Increase in reported infections per 100.000`)
  
   #create a plot of the increase per 100.000 inhabitants over time
   barplot(diff_frame$`Increase in reported infections per 100.000`, 
                  main = "Difference in reported infections per 100.000 inhabitants", 
                  xlab = "Date", 
                  ylab = "Difference in reported infections per 100.000 inhabitants", 
                  names.arg = as.Date(dates[-1]))

  return(diff_frame)
}

differenceCalculator2(used_dates)
```
```{r}
#in the data preparation, we already multplied the municipality population by 1000,
#so no need to do that again.
#however, we do need make a new variable containing the number of cases per 100.000 people
#to do that, we multiple the total number of reported cases by 100.000 divided by the municipality population and give it an informative name.
used_dates$cases_per_100000 <- used_dates$MN_total.reported * (100000/used_dates$MN_pop)

#make a function that calculates the mean number of infection per 100.000 inhabitants per day.
#so, this is over the Netherlands as a whole.
totalCalculator2 <- function(used_dates) {
  dates <- unique(used_dates$newdate) #create a list of dates
  totals <- numeric(length(dates)-1) #make an object with the length of one fewer than the number of dates. Why one fewer? We cannot calculate the number of new reported cases of the first day, because the number of cases from the preceding day is unknown.
  
  for (i in 1:(length(totals))) {
        step1 <- filter(used_dates, newdate == dates[i]) #select the date we want to calculate the nr of cases per 100.000 inhabitants for
    
    #create data frame of the  dates, and remove duplicates (total reported per municipalities are the same across duplicates)
    df <- step1[!duplicated(step1[,"MN_name"]),] 
    
    #calculate the mean number of reported cases per 100.000 inhabitants in the Netherlands and store in object
    means <- mean(df$cases_per_100000, na.rm = TRUE)
    totals[i] <- means #give it a name
  }
  
  #create a dataframe with the totals by the respecitve dates
  tot_frame <- as.data.frame(cbind(dates[-1], totals))
  #give the columns informative names
  colnames(tot_frame) <- c("Date", "Reported infections per 100.000")
  #and make the increase numeric
  tot_frame$`Reported infections per 100.000` <- as.numeric(tot_frame$`Reported infections per 100.000`)
  
   #create a plot of the increase per 100.000 inhabitants over time
   barplot(tot_frame$`Reported infections per 100.000`, 
                  main = "Mean number of reported infections per 100.000 inhabitants over time", 
                  xlab = "Date", 
                  ylab = "Mean number of reported infections per 100.000 inhabitants", 
                  names.arg = as.Date(dates[-1]))

  return(tot_frame)
}

totalCalculator2(used_dates)
```

#Extra: RNA particles.
```{r}
#make a function that calculates the RNA flow per 100.000 inhabitants per day in The Netherlands as a whole.
totalRNA <- function(used_dates) {
  dates <- unique(used_dates$newdate) #create a list of dates
  totals <- numeric(length(dates)-1) #make an object with the length of one fewer than the number of dates. Why one fewer? We cannot calculate the number of new reported cases of the first day, because the number of cases from the preceding day is unknown.
  
  for (i in 1:(length(totals))) {
    step1 <- filter(used_dates, newdate == dates[i]) #select the date we want to calculate the nr of cases per 100.000 inhabitants for
    
    #calculate the mean number of reported cases per 100.000 inhabitants in the Netherlands and store in object
    total <- sum(step1$TP_RNA.flow.per.100000, na.rm = TRUE) # Calculate the sum of all RNA measures on a given day
    totals[i] <- total #store it
  }
  
  #create a dataframe with the totals by the respecitve dates
  tot_frame <- as.data.frame(cbind(dates[-1], totals))
  #give the columns informative names
  colnames(tot_frame) <- c("Date", "RNA flow per 100.000")
  #and make the increase numeric
  tot_frame$`RNA flow per 100.000` <- as.numeric(tot_frame$`RNA flow per 100.000`)
  
   #create a plot of the increase per 100.000 inhabitants over time
   barplot(tot_frame$`RNA flow per 100.000`, 
                  main = "RNA flow per 100.000 inhabitants in The Netherlands over time", 
                  xlab = "Date", 
                  ylab = "RNA flow per 100.000 inhabitants", 
                  names.arg = as.Date(dates[-1]))

  return(tot_frame)
}

totalRNA(used_dates)
```

#Step 3: the mean the level of virus particles found in the water per 100.000 casess
```{r}

used_dates$TP_RNA.flow.per.100000 <- as.numeric(used_dates$TP_RNA.flow.per.100000)

virusCalculator <- function(used_dates) {
  dates <- unique(used_dates$newdate)
  means <- numeric(length(dates))
  for (i in 1:length(means)) {
    date_data <- filter(used_dates, newdate == dates[i])
    df1 <- date_data[!duplicated(date_data[,"MN_name"]),]
    means[i] <- mean(df1$TP_RNA.flow.per.100000, na.rm = TRUE)
  }
  means_frame <- as.data.frame(cbind(dates, means))
  colnames(means_frame) <- c("Date", "Mean RNA flow per 100000")
  means_frame$`Mean RNA flow per 100000` <- as.double(means_frame$`Mean RNA flow per 100000`)
  return(means_frame)
}

data_frame1 <- virusCalculator(used_dates)
```


#Step 4: mean number of infections per 100.000 inhabitants per day, weighted for the percentage of sewage water of a municipality by a treatment plant.

We believe that the value in the column 'TP_security.region' corresponds with the percentage of sewage water that a RWZI processes that comes from the security region it is in. Therefore, a sensible way of implementing the weighting is to take that percentage and multiply it with the RNA particle flow within that RWZI. We subsequently take the mean per day and see how it has affected the mean RNA flow.

### FOR OUR GROUP: I'VE REMOVED ALL NA ROWS FOR NOW. THIS WOULD HAVE ALSO BEEN DONE IF NA.RM HAD BEEN IMPLEMENTED IN CALCULATING THE MEAN

```{r}
rnaWeighting <- function(used_dates) {
  options(digits = 22)
  dates <- unique(used_dates$newdate) #create a list of all dates
  adjusted_means <- numeric(length(dates)) #create a vector of zeros that is equal in length to the number of dates
  
  for (i in 1:length(adjusted_means)) {
   non_na_data <- used_dates[!is.na(used_dates$TP_RNA.flow.per.100000),] #just for now eliminate NA rows
  non_na_data2 <- non_na_data[!is.na(non_na_data$TP_percentage.in.SR),]
  non_na_data2$TP_percentage.in.SR <- as.double(non_na_data2$TP_percentage.in.SR)
  non_na_data2$adjusted_RNA_flow <- (non_na_data2$TP_RNA.flow.per.100000 *    non_na_data2$TP_percentage.in.SR) #adjust rna flow by multiplying it with the proportion of water from that security region processed by that RWZI
     day_data <- filter(non_na_data2, newdate == dates[i])
    df1 <- day_data[!duplicated(day_data[,"MN_name"]),]
    
    
    adjusted_means[i] <- mean(df1$adjusted_RNA_flow, na.rm = TRUE)
  }
  means_frame <- as.data.frame(cbind(dates, adjusted_means))
  colnames(means_frame) <- c("Date", "Adjusted mean RNA flow per 100000")
  means_frame$`Adjusted mean RNA flow per 100000` <- as.double(means_frame$`Adjusted mean RNA flow per 100000`)
  
     #create a plot of the increase per 100.000 inhabitants over time
   barplot(means_frame$`Adjusted mean RNA flow per 100000`, 
                  main = "RNA flow per 100.000 inhabitants in The Netherlands over time", 
                  xlab = "Date", 
                  ylab = "RNA flow per 100.000 inhabitants", 
                  names.arg = as.Date(dates))
  
  return(means_frame)
}

data_frame2 <- rnaWeighting(used_dates)

mean(data_frame1$`Mean RNA flow per 100000`, na.rm = TRUE)
mean(data_frame2$`Adjusted mean RNA flow per 100000`, na.rm = TRUE)

```


#Step 5: how do the number of RNA particles per 100.000 inhabitants and the number of new cases relate to each other?

```{r}
sumRNAcalculator <- function(used_dates, date) {
  dates <- unique(used_dates$newdate) #create a list of dates
  means <- numeric(length(dates))
  step1 <- filter(used_dates, newdate == date)

  step1$MN_name <- as.factor(step1$MN_name)
  sum_rna_flows <- tapply(step1$TP_RNA.flow.per.100000, list(Municaplities = step1$MN_name), sum, na.rm = TRUE)

  return(sum_rna_flows)
}

totalReported <- function(used_dates, date) {
  step1 <- filter(used_dates, newdate == date)
  df1 <- step1[!duplicated(step1[,"MN_name"]),c(3,7)]
  
  return(df1)
}

differenceCalculator3 <- function(used_dates) {
  dates <- unique(used_dates$newdate) #create a list of dates
  cors <- numeric(length(dates)-7) #create a vector of zeros equal in length to the number of dates - 1 because we calculate differences
  cor_frame <- data.frame()
  
  for (i in 1:(length(cors))) {
    step1 <- filter(used_dates, newdate == dates[1]) #filter data from first day, e.g. 2020-09-07
    step2 <- filter(used_dates, newdate == dates[2]) #filter data from second day, e.g. 2020-09-08
    
    df1 <- step1[!duplicated(step1[,"MN_name"]),] #remove duplicate rows for total reported
    df2 <- step2[!duplicated(step2[,"MN_name"]),] #remove duplicate rows for total reported
    
    df1$MN_name <- as.factor(df1$MN_name) 
    step1$MN_name <- as.factor(step1$MN_name)
    step2$MN_name <- as.factor(step2$MN_name)
    ind1 <- sort.list(df1$MN_name)
    ind2 <- sort.list(df2$MN_name)
    
    diff_infections <- df2[ind2,]$MN_total.reported - df1[ind1,]$MN_total.reported #calculate difference in infections for each municipality
    
    diff_rna <- tapply(step2$TP_RNA.flow.per.100000, list(Municipalities = step2$MN_name), sum, na.rm = TRUE) - tapply(step1$TP_RNA.flow.per.100000, list(Municipalities = step1$MN_name), sum, na.rm = TRUE) #calculate difference in total rna flow for each municipality
    
    new_frame = data.frame(rep(dates[i+1], 344), df1$MN_name, diff_infections, diff_rna)
    colnames(new_frame) <- c("Date", "Municipality", "Change in infections", "Change in RNA")
    cor_frame <- rbind(cor_frame, new_frame)
  }

  return(cor_frame)
}

cor_frame <- differenceCalculator3(used_dates)
cor_frame$Municipality <- rownames(cor_frame)

means1day <- meanRNAcalculator(used_dates, "2020-09-07")
means1day[means1day == "NaN"] <- 0
means1day

infections <- totalReported(used_dates, "2020-09-14")
cor(means1day, infections$MN_total.reported)


means2day <- meanRNAcalculator(used_dates, "2020-10-20")
means2day[means1day == "NaN"] <- 0

infections2 <- totalReported(used_dates, "2020-10-27")
cor(means2day, infections2$MN_total.reported)

Utrecht <- used_dates[which(used_dates$MN_name=='Utrecht'), ]
Utrecht2 <- Utrecht[which(!is.na(Utrecht$TP_RNA.flow.per.100000)), ]

with(Utrecht2, cor(MN_total.reported, TP_RNA.flow.per.100000))
plot(Utrecht2$MN_total.reported, Utrecht2$TP_RNA.flow.per.100000)

sumRNAcalculator(Utrecht)
```

```{r}
sumRNAcalculator <- function(used_dates) {
  dates <- unique(used_dates$newdate) #create a list of dates
  sum_rna_flows <- numeric(length(dates))
  
  for (i in 1:length(sum_rna_flows)) {
  step1 <- filter(used_dates, newdate == dates[i])
  sum_rna_flows[i] <- sum(step1$TP_RNA.flow.per.100000, na.rm = TRUE)
  
  }

  sum_frame <- as.data.frame(cbind(dates, sum_rna_flows))
  colnames(sum_frame) <- c("Date", "sum_RNA")
  sum_frame$`sum_RNA` <- as.double(sum_frame$`sum_RNA`)
  
  return(sum_frame)
}

meanReportedcalculator <- function(used_dates) {
  dates <- unique(used_dates$newdate) #create a list of dates
  mean_reported <- numeric(length(dates))
  
  for (i in 1:length(mean_reported )) {
  step1 <- filter(used_dates, newdate == dates[i])
  mean_reported [i] <- mean(step1$MN_total.reported, na.rm = TRUE)
  
  }

  mean_frame <- as.data.frame(mean_reported)
  colnames(mean_frame) <- c("total_reported")
  mean_frame$`total_reported` <- as.double(mean_frame$`total_reported`)
  
  return(mean_frame)
}

Utrecht3 <- sumRNAcalculator(Utrecht2)
Utrecht4 <- meanReportedcalculator(Utrecht2)
Utrecht5 <- cbind(Utrecht3, Utrecht4)
cor(Utrecht5$total_reported, Utrecht5$sum_RNA)
plot(Utrecht5$total_reported, Utrecht5$sum_RNA)
abline(lm(total_reported ~ sum_RNA, data = Utrecht5))
```



```{r waste bin}
#because of this:
#multiply by 1000 to get rid of the "." in the population variable
#dat <- dat %>% mutate(MN_pop = MN_pop*1000)
#I think this line in step 2 is double, so I removed it and put it here
used_dates$MN_pop <- used_dates$MN_pop * 1000

```

