---
title: "Code for cleaning data"
author: "Lauke Stoel"
date: "26/11/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### DATA CLEANING

Rename column names to include a code for the level of measurement: 
SR = Security Region
MN = Municipality
TP = Treatment Plant

```{r}
dat <- dat %>% rename(SR_code = code, 
                        SR_name = naam,
                        MN_name = Municipality_name, 
                        MN_code = Municipality_code, 
                        MN_province = Province,
                        MN_total.reported = Total_reported, 
                        MN_hospital.admisstions = Hospital_admission, 
                        MN_deceased = Deceased, 
                        MN_pop = inhabitants, 
                        MN_popperkm2 = perkm.2, 
                        MN_sizekm2 = size,
                        TP_code = RWZI_AWZI_code, 
                        TP_name = RWZI_AWZI_name,
                        TP_Xco = X_coordinate,
                        TP_Yco = Y_coordinate, 
                        TP_postal.code = Postal_code, 
                        TP_security.region = Security_region_name, 
                        TP_percentage.in.SR = Percentage_in_security_region,
                        TP_RNA.per.ml = RNA_per_ml, 
                        TP_RNA.flow.per.100000 = RNA_flow_per_100000, 
                        TP_representative.measurement = Representative_measurement)
```

Re-order the data such that those of the same level of measurement are put together. The date_of_report variable has been left out - it's the same as newdate and newdate has a more convenient notation. 
```{r}
dat <- dat %>% select(newdate, SR_code, SR_name, MN_code, MN_name, MN_code:TP_RNA.per.ml, TP_RNA.flow.per.100000, TP_representative.measurement)
```

Structure of the dataset: 
```{r}
str(dat)
```
Some variables are reported as character strings, that should be numeric variables - such as MN_sizekm2, TP_percentage.in.SR and TP_RNA.flow.per.100000. This is because the Dutch notation for decimal numbers is comma's, where R would expect dots, and vice versa. Other variables are noted in the wrong order of size because of this. 
```{r}
#replace comma's by dots for all variables with decimals
dat$MN_sizekm2 <- gsub(",", ".", dat$MN_sizekm2)
dat$MN_popperkm2 <- gsub(",", ".", dat$MN_popperkm2)
dat$TP_percentage.in.SR <- gsub(",", ".", dat$TP_percentage.in.SR)
dat$TP_RNA.flow.per.100000 <- gsub(",", ".", dat$TP_RNA.flow.per.100000)

#multiply by 1000 to get rid of the "." in the population variable
dat <- dat %>% mutate(MN_pop = MN_pop*1000)

#redefine variables as numeric:
dat$TP_Yco <- as.numeric(dat$TP_Yco)
dat$MN_popperkm2 <- as.numeric(dat$MN_popperkm2)
dat$MN_sizekm2 <- as.numeric(dat$MN_sizekm2)
dat$TP_RNA.flow.per.100000 <- as.numeric(dat$TP_RNA.flow.per.100000)
dat$TP_percentage.in.SR <- as.numeric(dat$TP_percentage.in.SR)

#check the structure again
str(dat)
```
